A global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-02-26 Papers/Torque_Paper/article_IOP.tex
updates section 4-5
minor updates to paper
* edited conclusion
* changed x to r
* description of RVs
minor updates
- reducing text
- updating references
full paper draft, sent to ECN for check

% packages
%\usepackage{microtype}
%\usepackage{algorithm}
%\usepackage[noend]{algpseudocode}
%\usepackage[a4paper, total={6.2in, 9in}]{geometry}
%\renewcommand{\baselinestretch}{1.2} 
%\usepackage{color}
%\usepackage[x11names]{xcolor}
%\usepackage[utf8]{inputenc}
%\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsfonts}
%\usepackage{tikz}
% \usepackage{array}
\usepackage{mathtools}
\usepackage{relsize}
%\usepackage[T1]{fontenc}
\usepackage{cases}
\usepackage{amsthm}
\usepackage{subcaption}
\usepackage{multirow}
%\usepackage{stix2}
%\usepackage[labelfont=bf,font=footnotesize]{caption}
\usepackage{tabularx,ragged2e,booktabs}
\usepackage{todonotes}
%\usepackage{lineno}
%\modulolinenumbers[5]
%\usepackage[toc]{appendix}
\usepackage{hyperref}

@@

@@

\author{P.~Kumar$^{1}$, B.~Sanderse$^{1}$, K.~Boorsma$^{2}$, M.~Caboni$^{2}$} \ead{b.sanderse@cwi.nl}
\address{$^{1}$ Centrum Wiskunde \& Informatica (CWI), Amsterdam, The Netherlands}
\address{$^{2}$ TNO, Westerduinweg 3, Petten, The Netherlands}

@@

\begin{abstract}
A global sensitivity analysis of model parameters associated with the Blade Element Momentum (BEM) models is performed. The Sobol indices based on a sparse polynomial  expansion are used as a measure of global sensitivities. The sensitivity analysis workflow is developed using the uncertainty quantification toolbox \texttt{UQLab} that is integrated with the ECN's \texttt{Aero-Module} aeroelastic code. Sensitivity studies are performed on the NM80 wind turbine model from the DANAERO project.
\end{abstract}

%\begin{keyword}
%BEM, global sensitivity analysis, model uncertainty, Sobol indices
%\end{keyword}
%\end{frontmatter}

@@

%\newpage
\bibliography{\myreferences}

>>>>

% packages
%\usepackage{microtype}
%\usepackage{algorithm}
%\usepackage[noend]{algpseudocode}
%\usepackage[a4paper, total={6.2in, 9in}]{geometry}
%\renewcommand{\baselinestretch}{1.2} 
%\usepackage{color}
%\usepackage[x11names]{xcolor}
%\usepackage[utf8]{inputenc}
%\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsfonts}
%\usepackage{tikz}
% \usepackage{array}
\usepackage{mathtools}
\usepackage{relsize}
%\usepackage[T1]{fontenc}
\usepackage{cases}
\usepackage{amsthm}
\usepackage{subcaption}
\usepackage{multirow}
%\usepackage{stix2}
%\usepackage[labelfont=bf,font=footnotesize]{caption}
\usepackage{tabularx,ragged2e,booktabs}
\usepackage{todonotes}
%\usepackage{lineno}
%\modulolinenumbers[5]
%\usepackage[toc]{appendix}
\usepackage{placeins}
\usepackage{hyperref}

@@

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\author{P.~Kumar$^{1}$, B.~Sanderse$^{1}$, K.~Boorsma$ ^{2}$, M.~Caboni$^{2}$} \ead{b.sanderse@cwi.nl}
\address{$^{1}$ Centrum Wiskunde \& Informatica (CWI), Amsterdam, The Netherlands}
\address{$^{2}$ TNO, Westerduinweg 3, Petten, The Netherlands}

@@

\begin{abstract}
A framework is presented for performing global sensitivity analysis of model parameters associated with the Blade Element Momentum (BEM) models. The Sobol indices based on a sparse polynomial  expansion are used as a measure of global sensitivities. The sensitivity analysis workflow is developed using the uncertainty quantification toolbox \texttt{UQLab} that is integrated with the ECN's \texttt{Aero-Module} aeroelastic code. Sensitivity studies are performed on the NM80 wind turbine model from the DANAERO project.
\end{abstract}

\vspace{-1cm}
%\begin{keyword}
%BEM, global sensitivity analysis, model uncertainty, Sobol indices
%\end{keyword}
%\end{frontmatter}

@@

%\newpage
\section*{References}
\bibliography{\myreferences}

### 2020-02-26 Papers/Torque_Paper/content.tex
full paper draft, sent to ECN for check

Recently, several papers have addressed the uncertainty in BEM model output by performing uncertainty propagation and sensitivity studies, e.g.\ \cite{Echeverria2017,Matthaus2017,Murcia2018,Robertson2018,Bos2019a}. In these studies, the focus in mainly on uncertainties in the external conditions (wind parameters) and/or uncertainty in the turbine specification (geometric parameters). Apart from understanding how uncertainty in the output is related to the different uncertain inputs, such sensitivity studies are very useful to reduce the number of parameters as needed for example in design optimization \cite{Echeverria2017}. However, the uncertainty in BEM model output as caused by uncertainty in the model formulation itself, e.g.\ through the values chosen for model parameters, has been given little attention (exceptions being the effect of aerodynamic properties studied in \cite{Bortolotti2019,Matthaus2017})\todo{this is also what we do}.

The goal of this work is to perform a systematic assessment of the uncertainty in model parameters in BEM models. We approach this by performing a global sensitivity analysis based on the Sobol expansion approach, which decomposes the total variance of the quantity of interest (model output) into contributions from individual parameters and their combinations, similar to \cite{Echeverria2017,Murcia2018, Rinker2016a}. We will employ the uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}, which computes the Sobol indices based on a sparse polynomial chaos expansion. The main novelty of the work lies in creating a generic framework for studying parametric uncertainteis, including both model uncertainties and geometric uncertainties. As such, the current sensitivity analysis is part of the so-called WindTrue project, in which the long-term goal is the development of calibrated BEM models. Anticipating on this goal, we are using the NM80 wind turbine model from the Tjaereborg wind farm, for which extensive measurement data is available through the DanAero project \cite{Troldborg2013} and IEA Task 29. The current sensitivity analysis will make clear which of the model parameters are to be used in the calibration phase.

@@

\section{Aeroelastic model uncertainties}\label{sec:model_description}
%\subsection{BEM model uncertainties}
State-of-the-art wind-turbine models simulate the aeroelastic behavior of wind turbines by combining the concept of momentum conservation of the flow (BEM theory) with the equations of motion for the structure, possibly extended with the hydrodynamics of the sea and control algorithms \cite{Vorpahl2013}. In this work, we concentrate on the first aspect, namely the prediction of flow and blade forces as given by the BEM method. The particular BEM code that we use is ECN's \texttt{Aero-Module} \cite{Boorsma2012}. For the case of a rigid turbine, with a uniform inflow field, the main uncertainties that arise in the BEM formulation are \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurements or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modeled via \textit{a dynamic stall model} and associated time constant;
\item Yawed inflow is not naturally included and typically incorporated via \textit{a wake correction}.
%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model
\end{itemize}
Next to these model uncertainties we will consider geometric uncertainties: the chord and twist distribution along the blade.

@@

There are two ways to obtain perturbed curves: independently perturbing each of the control points (local perturbation) or using a single random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch3 - Ch7 and twist control points Tw2 - Tw7 control points, see Fig.\ \ref{perturbed_samples} (a) - (b), respectively. The twist and chord are perturbed following equation \eqref{eqn:perturb_c}, whereas the lift and drag are perturbed according to equation \eqref{eqn:perturb_cl}. Examples of the resulting lift and drag curves are shown in Fig.\ \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this purpose. Also note that when using a uniform distribution with bounds given by the relative magnitude of control points, we tend to obtain small perturbations for control points with values close to zero. 
\begin{figure}[h!]
\centering
\includegraphics[trim={0.5cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

@@

@@

\section{Conclusions}\label{sec:conclusions}
We have shown how Sobol indices computed using sparse adaptive polynomial chaos expansion can be used for high-dimensional global sensitivity analysis of both geometric and model uncertainties in BEM models. The proposed framework is generic and can deal with many types of (parametric uncertainties) The sensitive model parameters identified here will be utilized in future work to develop calibrated BEM models with built-in uncertainty estimates, as part of the \textit{WindTrue} project, and used to better understand the differences between BEM models and measurement data as studied in IEA Task 29.

>>>>

Recently, several papers have addressed the uncertainty in BEM model output by performing uncertainty propagation and sensitivity studies, e.g.\ \cite{Echeverria2017,Matthaus2017,Murcia2018,Robertson2018,Bos2019c}. In these studies, the focus in mainly on uncertainties in the external conditions (wind parameters) and/or uncertainty in the turbine specification (geometric parameters). Apart from understanding how uncertainty in the output is related to the different uncertain inputs, such sensitivity studies are very useful to reduce the number of parameters as needed for example in design optimization \cite{Echeverria2017}. However, the uncertainty in BEM model output as caused by uncertainty in the model formulation itself, e.g.\ through the values chosen for model parameters, has been given little attention. % (exceptions being the effect of aerodynamic properties studied in \cite{Bortolotti2019,Matthaus2017})\todo{this is also what we do}.

The goal of this work is to perform a systematic assessment of the uncertainty in model parameters in BEM models. We approach this by performing a global sensitivity analysis based on the Sobol expansion approach, which decomposes the total variance of the quantity of interest (model output) into contributions from individual parameters and their combinations, similar to \cite{Echeverria2017,Murcia2018, Rinker2016a}. We will employ the uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}, which computes the Sobol indices based on a sparse polynomial chaos expansion. The main novelty of the work lies in creating a generic framework for studying parametric uncertainteis, including both model uncertainties and geometric uncertainties. This framework for sensitivity analysis is part of the so-called \textit{WindTrue} project, in which the long-term goal is the development of calibrated BEM models. Anticipating on this goal, we are using the NM80 wind turbine model from the Tjaereborg wind farm, for which extensive measurement data is available through the DanAero project \cite{Troldborg2013} and IEA Task 29. The sensitivity analysis will make clear which of the model parameters are to be used in the calibration phase.

@@

\section{Aeroelastic model uncertainties}\label{sec:model_description}
%\subsection{BEM model uncertainties}
State-of-the-art wind-turbine models simulate the aeroelastic behavior of wind turbines by combining the concept of momentum conservation of the flow (BEM theory) with the equations of motion for the structure, possibly extended with the hydrodynamics of the sea and control algorithms \cite{Vorpahl2013}. In this work, we concentrate on the first aspect, namely the prediction of flow and blade forces as given by the BEM method. The particular BEM code that we use is ECN's \texttt{Aero-Module} \cite{Boorsma2012}. For the case of a rigid turbine, with a uniform inflow field, the main uncertainties that arise in the BEM formulation are \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurements or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modeled via \textit{a dynamic stall model} and associated time constant;
\item Yawed inflow is not naturally included and typically incorporated via \textit{a wake correction}.
%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model
\end{itemize}
In this paper we will focus on the first uncertainty -- uncertainty in lift and drag coefficients -- but we stress that the framework and methodology are directly applicable to other BEM model uncertainties as well. In a parallel paper, see \cite{Caboni2020}, the effect of other uncertainties such as the dynamic stall model will be investigated. Next to these model uncertainties we will consider geometric uncertainties: the chord and twist distribution along the blade.

@@

There are two ways to obtain perturbed curves: independently perturbing each of the control points (local perturbation) or using a single random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch3 - Ch7 and twist control points Tw2 - Tw7 control points, see Fig.\ \ref{perturbed_samples} (a) - (b), respectively. The twist and chord are perturbed following equation \eqref{eqn:perturb_c}, whereas the lift and drag are perturbed according to equation \eqref{eqn:perturb_cl}. Examples of the resulting lift and drag curves are shown in Fig.\ \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this purpose. Also note that when using a uniform distribution with bounds given by the relative magnitude of control points, we tend to obtain small perturbations for control points with values close to zero. 
\begin{figure}[ht!]
\centering
\includegraphics[trim={0.5cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

@@

%\FloatBarrier

@@

\FloatBarrier

\section{Conclusions}\label{sec:conclusions}
In this paper we have developed a framework for global sensitivity analysis using Sobol indices computed using sparse adaptive polynomial chaos expansions. The framework can be used for high-dimensional global sensitivity analysis of both geometric and model uncertainties in BEM models. As such, the proposed framework is generic and can deal with many types of (parametric) uncertainties, of which a few were investigated in this paper: twist, chord, lift coefficient and drag coefficient. By using NURBS parameterizations of chord, twist, lift and drag as a function of the blade radius, the number of uncertainties could be limited to 19. The results highlight amongst others the importance of the lift coefficient, especially for the axial force prediction. The identified sensitivities will be utilized in future work to develop calibrated BEM models with built-in uncertainty estimates, as part of the \textit{WindTrue} project, and used to better understand the differences between BEM models and measurement data as studied in IEA Task 29.
